---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r}
knitr::opts_chunk$set(echo=TRUE)
library(tidyr)
library(dplyr)
library(ggplot2)
```

## Load and preprocess the data
The date dimension was originally a 'char'. Best to change this to 'Date'.

```{r}
activity <- read.csv("activity.csv")
str(activity)
activity$date <- as.Date(activity$date)
str(activity)
```

## What is mean total number of steps taken per day?
First, I needed to get the steps per day. Then I was able to get to the mean total number of steps per day.

```{r}
steps_per_day <- activity %>% group_by(date) %>% 
  summarize(steps_sum=sum(steps,rm.na=TRUE))
ggplot(steps_per_day,aes(x=steps_sum)) + geom_histogram() + 
  labs(title='Histogram of Total Steps Per Day',x='Steps',y='Frequency Count')

print(paste('Median total steps per day:',median(steps_per_day$steps_sum, 
  na.rm = TRUE)))
print(paste('Mean total steps per day:',as.integer(mean(steps_per_day$steps_sum, 
  na.rm = TRUE))))
```

## What is the average daily activity pattern?
I determined the average steps for each interval and then graphed it over time. Then I mapped the max back to the interval to highlight the most active interval on average.

```{r}
steps_per_interval <- activity %>% group_by(interval) %>% 
  summarize(steps_mean=mean(steps,na.rm = TRUE))
ggplot(steps_per_interval,aes(x=interval,y=steps_mean)) + geom_line() + 
  labs(title='Average Daily Activity',x='Interval',y='Step Mean')
max_interval <- steps_per_interval[steps_per_interval$steps_mean == 
                                     max(steps_per_interval$steps_mean),]$interval
print(paste('Interval with most steps on average:',max_interval))
```

## Imputing missing values
I determined there were missing values for steps. I then replaced those with the interval average. I replotted the histogram and calculated the median and mean to see if there was much of a difference, which there wasn't.

```{r}
summary(activity)
print(paste("Total # of rows with missing NAs, which only occurs with 'steps': ",
  sum(!complete.cases(activity$steps))))
activity_cleansed <- activity %>% group_by(interval) %>% mutate(steps = 
  replace_na(steps, mean(steps, na.rm = TRUE)))
steps_per_day_cleansed <- activity_cleansed %>% group_by(date) %>% 
  summarize(steps_sum=sum(steps))

ggplot(steps_per_day_cleansed,aes(x=steps_sum)) + geom_histogram() + 
  labs(title='Histogram of Total Steps Per Day, Cleansed',x='Steps',
       y='Frequency Count')

print(paste('Median total steps per day:',
  as.integer(median(steps_per_day_cleansed$steps_sum, na.rm = TRUE))))
print(paste('Mean total steps per day:',
  as.integer(mean(steps_per_day_cleansed$steps_sum, na.rm = TRUE))))
```

## Are there differences in activity patterns between weekdays and weekends?
First, I created a function to determine which days were weekdays or weekends. Then I did a panel plot, plotting the two side by side. Weekends have more activity throughout the day, as expected.
```{r}
wd_or_we <- function(x) {
  day_name<-weekdays(x)
  if(day_name=='Saturday' | day_name=='Sunday'){
    'weekend'
  } else {
    'weekday'
  }
}
activity_cleansed$day_type <- sapply(activity_cleansed$date,wd_or_we)
head(activity_cleansed)

to_plot <- activity_cleansed %>% group_by(day_type,interval) %>% 
  summarize(avg_steps=mean(steps))
ggplot(to_plot,aes(x=interval,y=avg_steps)) + geom_line() + 
  facet_grid(day_type ~ .) + labs(title='Average Daily Activity',x='Interval',
                                  y='Step Mean')
```